---
title: "Estimating pi (Quarto method)"
author: "Stephen Eglen"
date: today
bibliography: refs.bib
---

This is an example document created using the quarto system
<https://quarto.org>.  For this document, the master file is
`estimateq.qmd`.  This is processed by quarto using the following
shell command

```{bash}
#| eval: false
quarto render estimateq.qmd
```

Within RStudio, there is a handy "Knit HTML"  button.


# Preamble


```{r setup,collapse=TRUE}
require(knitr) ## for kable
```

TODO: does quarto allow for caching?

#  Task: estimate the value of $\pi$

Our task is to estimate the value of $\pi$ by simulating darts being
thrown at a dartboard.  Imagine that the person throwing the darts is
not very good, and randomly throws each dart so that it falls
uniformly within a square of side length $2r$, with the dartboard of
radius $r$ centred within that square.  If the player throws $n$
darts, and $d$ of them hit the dartboard, then for large enough $n$,
the ratio $d/n$ should approximate the ratio of the area of the
dartboard to the enclosing square, $\pi r^2 / 4 r^2 \equiv \pi/4$.
From this, we can estimate $\pi \approx 4d/n$.

We start with an example, shown in @fig-dartboard, using R to draw both the dartboard and the
surrounding square, together with `n=50` darts.  The radius of the
dartboard here is 1 unit, although the value is not important.

```{r}
#| label: fig-dartboard
#| fig-cap: example dartboard.
r <- 1
n <- 50
par(las=1)
plot(NA, xlim=c(-r,r), ylim=c(-r,r), asp=1, bty='n',
     xaxt='n', yaxt='n', xlab='', ylab='')
axis(1, at=c(-r,0,r)); axis(2, at=c(-r,0,r))
symbols(x=0, y=0, circles=r, inch=F, add=T)
x <- runif(n, -r, r);  y <- runif(n, -r, r)
inside <- (x^2 + y^2) < r^2
d <- sum(inside)
points(x, y, pch=ifelse(inside, 19, 4))
rect(-r, -r, r, r, border='blue', lwd=2)
```

A dart is drawn as a filled circle if it falls within the dartboard,
else it is drawn as a cross.  In this case the number of darts within
the circle is `r d`, and so the estimated value is $\pi \approx
`r {4*d/n}`$.

The estimate of $\pi$ should improve as we increase the number of darts
thrown at the dartboard.  To verify this, we write a short function
that, given the number of darts to throw, $n$, returns an estimate of
$\pi$.

```{r}
estimate.pi <- function(n=1000) {
  ## Return an estimate of PI using dartboard 
  ## method with N trials.
  r <- 1 ## radius of dartboard
  x <- runif(n, min=-r, max=r)
  y <- runif(n, min=-r, max=r)
  l <- sqrt(x^2 + y^2)
  d <- sum(l<r)
  4*d/n
}
```

We can then test the procedure a few times, using the default number
of darts, 1000:

```{r}
replicate(9, estimate.pi())
```
Finally, for a given value of $n$, we can show 99 estimates of $\pi$,
as clearly the estimate will vary from run to run.  In @fig-estimates we compare the estimates of $\pi$ for three
different values of $n$.


```{r}
#| fig-cap: Estimates of Pi as we increase number of darts
#| label: fig-estimates
ns <- 10^c(2,3,4)
res <- lapply(ns, function(n) replicate(99, estimate.pi(n)))
par(las=1, bty='n')
stripchart(res, method="jitter", group.names=ns, 
           xlab="number of darts",
           ylab=expression(paste('estimate of ', pi)),
           vert=TRUE, pch=20, cex=0.5)
abline(h=pi, col='red')
```

# Example table output

Here is a simple table.

```{r}
df = data.frame(name=c("joe", "ann", "bob"),
  age=c(19, 24, 27),
  height=c(1.8, 1.75, 1.7))
df
```

Or you can add a caption as shown in @tbl-example:

```{r}
#| label: tbl-example
#| echo: false
#| tbl-cap: Example output from kable function (part of knitr)
kable(df, format="markdown")
```

[Further information on tables](https://quarto.org/docs/authoring/tables.html#computations)

# References to literature

Bookdown can refer to references.  For example, this is the original
citation for R [@Ihaka1996].  For more details regarding knitr, check
out the book [@knitrbook].


[Further information on citations](https://quarto.org/docs/authoring/footnotes-and-citations.html)

# Exercises

1. Add an extra set of trials for n=10 darts thrown.

2. Experiment switching the cache on (Section \@ref(caching)).  Where
   are the intermediate results stored?  When is the code rerun?

3. Add another reference to the document.  (BibTeX entries can be
   downloaded from Google Scholar, amongst other places.)

4. Can you get pdf output as well as HTML output?  (Hint: check Makefile)

5. Explore the printr package for printing tables and matrices <https://yihui.name/printr/>.


## Compiling this document

```{r eval=FALSE}
quarto render estimateq.qmd
quarto_render("estimateq.qmd", output_format="pdf")
```

# References
